---
title: \sf MEDS5420 - Xanadu
header-includes:
- \usepackage{color}
- \usepackage{float}
- \DeclareUnicodeCharacter{2212}{-}
date: "February 4, 2026"
output:
  bookdown::html_document2:
    toc: true
fontsize: 14pt
geometry: margin=1in
---

# A review of `wget`, `curl`, and `cut` 

I asked you to complete section 5.6 from Lecture 4 for homework, so let's review these commands.

## Retrieving files from a URL using `curl` or `wget` 

Download `chr_coordinates.bed` from the Lecture 5 folder in GitHub and move this file to your MEDS5420 folder.  \

If you have a Mac, use `curl`:
```{r engine='bash', eval=F, echo=T}
curl -O https://path/to/the/raw/file/on/github.bed 
```

Linux OS have `wget`:

```{r engine='bash', eval=F, echo=T}
wget https://path/to/the/raw/file/on/github.bed
```

## Move the `bed` file to a new directory

```{r engine='bash', eval=F, echo=T}
#start from the MEDS5420 folder:
mkdir ./in_class/coordinates
mv chr_coordinates.bed ./in_class/coordinates
cd ./in_class/coordinates
```

## String splitting and manipulation with `cut`

Review the `cut` command by splitting the file name `chr_coordinates.bed` on the `r` character and then the `c` character to output `oo`. Remember to use `echo` to interpret the file name as a string and not a file.

```{r engine='bash', eval=T, echo=T}
fileName="chr_coordinates.bed"
echo $fileName | cut -f 2 -d 'r' | cut -f 2 -d 'c' 
```

# Logical tests in `Shell`
We've learn some basics of shell scripting with loops.  Let's add more sophistication by adding conditional statements.

## Example: **`if/else`**

```{r engine='bash', eval=T, echo=TRUE}
snow="1 2 3 4 6 8 10"
for x in $snow
  do
    echo "There are ${x} inches of snow"
    if [ $x -lt 3 ]
      then
      echo 'stay calm'
    else
      echo 'panic'
    fi
  done
```
*Note: the test statement must be separated from the square brackets by a space.\
End `if` loops with `fi`\
\

To get a list of operators for numerical tests use:
```{r engine='bash', eval=F, echo=TRUE}
man test
```

What if I want to add a layer(s) of contingency here: `elif`
```{r engine='bash', eval=T, echo=TRUE}
snow="1 2 3 4 6 8 10"
type=windy
for x in $snow
  do
    echo "There are ${x} inches of snow"
    if [ $x -lt 3 ]
      then
      echo 'stay calm'
    elif [ $x -lt 8 ] && [ ${type} = windy ]
      then
      echo 'it is windy, take cover'
    else
      echo 'ignore the wind and grab your sled'
    fi
  done
```

Note that I also included a variable that is used for interpretation.
\
\
`&&` represents *and* creating an **if/and** statement \
`||` represents *or* creating an **if/or** statement

# Passing files and options into scripts from command line:

I uploaded a file to the Lecture 6 directory in GitHub with rain data and I want to process with this script.
```{r engine='bash', eval=T, echo=TRUE}
cat rain_data.txt
```
I could read it in directly:
```{r engine='bash', eval=T, echo=TRUE}
rain=$(cat rain_data.txt | cut -f 2)
cond=windy
for x in $rain
  do
    echo "There are ${x} inches of rain"
    if [ $x -lt 3 ]
      then
      echo 'stay calm'
    elif [ $x -gt 5 ] && [ ${cond} == windy ]
      then
      echo 'it is windy, take cover'
    else
      echo 'get in a boat'
    fi
  done
```
Or, I could set a variable as below and save this as **`rain.sh`** script.

```{r engine='bash', eval=F, echo=TRUE}
#! /usr/bin/sh

rain=$(cat "$1" | cut -f "$2")
cond=$3
for x in $rain
  do
    echo "There are ${x} inches of rain"
    if [ $x -lt 3 ]
      then
      echo 'stay calm'
    elif [ $x -gt 5 ] && [ $cond == windy ]
      then
      echo 'it is windy, take cover'
    else
      echo 'get in a boat'
    fi
  done
```
**Note:** the `$1` usage here is a shortcut that allows the user to add an input file in the first argument.  The usage would then be: `<script_name> ARG1 ARG2 ARG3 ` \
`$1` refers to `rain_data.txt` \
`$2` refers to the number `2`, which happens to be our second argument and in the script it is used to parse out the second column \
`$3` refers to the condition, which is the third argument


Then I would run:
```{r engine='bash', eval=F, echo=TRUE}
# script input1 input2
bash rain.sh rain_data.txt 2 calm
# OR
bash rain.sh rain_data.txt 2 windy
# OR
chmod +x rain.sh
./rain.sh rain_data.txt 2 windy

```

More arguments can be added and the order of the arguments sets the substitution order.

## Reviewing how to pass shell variables to awk:

**Question:** What if we have a shell variable that we want to use or pass to **awk**?

Try this:
```{r engine='bash', eval=F, echo=T, error=F}
list="1 2 3"
echo $list | awk '{print $list}'
```
\
It doesn't work because a variable made in the shell cannot inherently be read by `awk`. You have to pass the variable to `awk`.  Here's how:
```{r engine='bash', eval=T, echo=T}
list="1 2 3"
echo $list| awk -v nums="$list" '{print nums}'
```

Recall the `-v` option in the beginning of the awk command.\


## In class exercise 1:
In the class last week we used `color-table.txt` and learned how to isolate and parse different columns and rows with **cut, uniq** and **awk**.  Now, try writing a script that will use the color column to parse each row to a file with identical colors only.  That is, all the 'red' rows should go to 'red.txt' file, blue; to a 'blue.txt' file, etc.

### Proper script structure and annotation
Even though your code may have worked, the script is not considered finished as it stands.  We need to use indentation and add annotation to the code for several reasons.\
**1.** Proper indentation of loop makes the code more readable. \
**2.** To provide USAGE instructions \
**3.** To describe the steps being taken.  This is important to remind yourself what your coding steps were or for other that might want to modify your script. \
**4.** To track the progress of the script.  This is most important for debugging, so that one can know where in the code a script failed. \

**Editing in a text editor with syntax highlighting will help construct a readable script**
\pagebreak


# Logging into the Xanadu cluster 

To access the cluster you need to login with **ssh** (secure shell):

```{r engine='bash', eval=F, echo=TRUE}
ssh <user_name>@xanadu-submit-ext.cam.uchc.edu 

# you user name looks like this:
ssh meds5420usr17@xanadu-submit-ext.cam.uchc.edu 
```
Once you login you are in the head (or login) node. **DO NOT** run any resource intensive commands on the head node. We will go over the procedure for allocating resources using both interactive sessions and job submissions.  

# Tranferring data between to and from cluster


## For PC users:
sdYou can use scp from a terminal window, or you can use `WinSCP` which is a convenient FTP client or user interface for transferring data between computers.  Below are some links with tutorials for downloading, installing, and using `WinSCP`.

[https://winscp.net/eng/docs/guide_connect](https://winscp.net/eng/docs/guide_connect) \

[https://www.youtube.com/watch?v=58KmUBaEW34](https://www.youtube.com/watch?v=58KmUBaEW34)


To move files in between computer you can login with **sftp** use **scp** (secure copy):

## **`sftp`**: 

`ftp` stands for "File Transfer Protocol", `sftp` is " Secure File Transfer Protocol".  In other words, with sftp, a useraccount and password are required.\
```{r engine='bash', eval=F, echo=TRUE}

sftp  <your_username>@<host_name>
```
For the Xanadu cluster, there is a special partition for transferring data:

```{r engine='bash', eval=F, echo=TRUE}

sftp <your_username>@transfer.cam.uchc.edu
```

**1.** You can then navigate to the directory where you want to take files from.\
**2.** `put` and `get` can be used to move files from or to your computer, respectively \

```{r engine='bash', eval=F, echo=TRUE}
# copy a document to the cluster
put /Users/guertinlab/MEDS5420/color-table.txt

# retrieve a copy of a document from the cluster (will go in whatever folder you logged in from)
get /home/FCAM/meds5420/usr17/file.txt


```


## **`scp`**
**scp** can be used without logging in provided you know the exact location where your file of interest is or will go. I find `sftp` easier and will use `sftp` for class.

```{r engine='bash', eval=F, echo=TRUE}
# for copying TO the server
scp -r <path_to_directory> <your_username>@transfer.cam.uchc.edu:~/path/to/target/folder
```

You should be prompted for a password.  If not, the transfer probably failed.
```{r engine='bash', eval=F, echo=TRUE}
# for copying FROM the server
scp -r <your_username>@<host_name>:@transfer.cam.uchc.edu <target_directory> 
```

# Logging into the Xanadu cluster 

To access the cluster you need to login with **ssh** (secure shell):

```{r engine='bash', eval=F, echo=TRUE}
ssh <user_name>@xanadu-submit-ext.cam.uchc.edu 

# you user name looks like this:
ssh meds5420usr17@xanadu-submit-ext.cam.uchc.edu 
```

# Tranferring data to and from the cluster

## For PC users:
You can use scp from a terminal window, or you can use `WinSCP` which is a convenient FTP client or user interface for transferring data between computers.  Below are some links with tutorials for downloading, installing, and using `WinSCP`.

[https://winscp.net/eng/docs/guide_connect](https://winscp.net/eng/docs/guide_connect) \

[https://www.youtube.com/watch?v=58KmUBaEW34](https://www.youtube.com/watch?v=58KmUBaEW34)


To move files in between computer you can login with **sftp** use **scp** (secure copy):

## **`sftp`**: 

`ftp` stands for "File Transfer Protocol", `sftp` is " Secure File Transfer Protocol".  In other words, with sftp, a useraccount and password are required.\
```{r engine='bash', eval=F, echo=TRUE}

sftp  <your_username>@<host_name>
```
For the Xanadu cluster, there is a special partition for transferring data:

```{r engine='bash', eval=F, echo=TRUE}

sftp <your_username>@transfer.cam.uchc.edu
```

**1.** You can then navigate to the directory where you want to take files from.\
**2.** `put` and `get` can be used to move files from or to your computer, respectively \


```{r engine='bash', eval=F, echo=TRUE}
put /Users/guertinlab/MEDS5420/color-table.txt

get 
```


## **`scp`**
**scp** can be used without logging in provided you know the exact location where your file of interest is or will go. We will primarily use `sftp` in this course.

```{r engine='bash', eval=F, echo=TRUE}
# for copying TO the server
scp -r <path_to_directory> <your_username>@transfer.cam.uchc.edu:~/path/to/target/folder
```

You should be prompted for a password.  If not, the transfer probably failed.
```{r engine='bash', eval=F, echo=TRUE}
# for copying FROM the server
scp -r <your_username>@<host_name>:<target_directory> 
```

### How do we know if transfer was complete?

There's a program called `md5` (mac) or `md5sum` (linux) that can help us with this.  It returns a compact digital fingerprint for each file.  Any change to the file will result in a different fingerprint.

on a mac:
```{r engine='bash', eval=T, echo=TRUE}
md5 ./data-shell.tar
```

on Linux:
```{r engine='bash', eval=F, echo=TRUE}
md5sum ./data-shell.tar
```

## In class exercise 2: Inspecting, retrieving, and checking files from server

**1** Log onto the server using ssh \
**2** Navigate to the MEDS5420 folder in `/home/FCAM/meds5420/in_class` \
**3** View the contents of the `data-shell.tar` file without unbundling it. \
**4** View the checksum string for the file. \
**5** Logout and return to your home directory or open a new terminal window (command-t) \
**6** Transfer the file to your computer using sftp \
**7** Confirm that the transfer was complete \


\pagebreak

# Answers to in class exercises:

## In class exercise 1:
```{r engine='bash', eval=F, echo=T}

colors=$(cat "$1"| cut -f 3 | sort | uniq)

for col in $colors
	do	
		touch ${col}_rows.txt
		cat "$1" | awk -v col="$col" '{ if ($3 == col) print $0}' >> ${col}_rows.txt
	done
```

Here's another version of the script with decent annotation. The input file would be the color-table.txt file. The first argument then replaces the "$1" wherever it appears in the script.  

```{r engine='bash', eval=F, echo=T}
# This script will parse unique items from column 3 to separate files
# USAGE:bash parse_colors.sh <INPUT_FILE>

#Create uniq list of colors.  Note that input file is first argument
colors=$(cat "$1"| cut -f 3 | sort | uniq)
echo $colors

#iterate through list of colors and parse into new files
for col in $colors
	do	
		echo parsing ${col} # this prints the progress to the screen
		touch ${col}_rows.txt
		cat "$1" | awk -v col="$col" '{ if ($3 == col) print $0}' >> ${col}_rows.txt
		echo ${col} parsed	# this prints the progress to the screen
	done
```

This version requires a Google search to figure out how to pass multiple variables to `awk`. This script is general and you can use it to parse any file based on unique elements in any user-defined column. Maybe whether genes are activated, unchanged, or repressed is in column 7, can we use this script to get a file for each gene category?

```{r engine='bash', eval=F, echo=T}

#! /bin/sh
# positional argument 1 is the file with the color data
# positional argument 2 is the column with the color 
color=$(cat $1 | cut -f $2 | sort | uniq)
for x in $color
  do
    echo "parsing the ${x} file"
    cat $1 | awk -v col="$x" -v fileName="$1" -v column="$2" '{if ($column == col) print $0}' > ${x}.txt
  done
```





## In class exercise 2:

### ssh 
```{r engine='bash', eval=F, echo=T}
ssh meds5420usr17@xanadu-submit-ext.cam.uchc.edu
#the 17 refers to your user number
cd /home/FCAM/meds5420/in_class
tar -tvf data-shell.tar
md5sum data-shell.tar
# a174bf3795d25f39891f43571ba1c678 
exit
```
Note that none of these commands demand significant compute resources.

### sftp
```{r engine='bash', eval=F, echo=T}
cd ~
sftp meds5420usr17@transfer.cam.uchc.edu
cd /home/FCAM/meds5420/in_class
get data-shell.tar
exit
md5 data-shell.tar # mac
#OR
md5sum data-shell.tar # linux
```


