---
title: \sf MEDS5420 - awk
header-includes:
- \usepackage{color}
- \usepackage{float}
- \DeclareUnicodeCharacter{2212}{-}
date: "February 4, 2026"
output:
  bookdown::html_document2:
    toc: true
fontsize: 14pt
geometry: margin=1in
---
# A review of `wget`, `curl`, and `cut`

I asked you to complete section 5.6 from Lecture 4 for homework, so let's review these commands.

## Retrieving files from a URL using `curl` or `wget`

Download `chr_coordinates.bed` from the Lecture 5 folder in GitHub and move this file to your MEDS5420 folder.  \

If you have a Mac, use `curl`:
```{r engine='bash', eval=F, echo=T}
curl -O https://path/to/the/raw/file/on/github.bed
```

Linux OS have `wget`:

```{r engine='bash', eval=F, echo=T}
wget https://path/to/the/raw/file/on/github.bed
```

## Move the `bed` file to a new directory

```{r engine='bash', eval=F, echo=T}
#start from the MEDS5420 folder:
mkdir ./in_class/coordinates
mv chr_coordinates.bed ./in_class/coordinates
cd ./in_class/coordinates
```

## String splitting and manipulation with `cut`

Review the `cut` command by splitting the file name `chr_coordinates.bed` on the `r` character and then the `c` character to output `oo`. Remember to use `echo` to interpret the file name as a string and not a file.

```{r engine='bash', eval=T, echo=T}
fileName="chr_coordinates.bed"
echo $fileName | cut -f 2 -d 'r' | cut -f 2 -d 'c' 
```
